---
title: "Distributions Boxplots"
subtitle: "Exercise 1"
author: "Javier Fong - 1004379994"
output: 
  html_document:
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 4
    theme: flatly
---

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(reshape2)
```

# Theoretical values

Obtain the five quantities used to define a boxplot for the following  distributions:

* Cauchy.
* Chi-square with 1 degree of freedom.
* Exponential distribution with $\lambda = 1$
* F-distribution with 5 and 2 degrees of freedom. 
* Student-t distribution with 5 degrees of freedom. 

```{r}
relevant_q = c(0.25, 0.5, 0.75)

boxplot_values = data.frame(
  cauchy = qcauchy(relevant_q)
  , chisqr = qchisq(relevant_q, df = 1)
  , exp = qexp(relevant_q, rate = 1)
  , fdist = qf(relevant_q, df1 = 5, df2 = 2)
  , student = qt(relevant_q, df = 5)
) %>% 
  t() %>% 
  data.frame() %>% 
  dplyr::rename(first_q = X1, second_q = X2, third_q = X3) %>% 
  dplyr::mutate(
    iqr = third_q - first_q
    , lower_b = first_q - (1.5 * iqr)
    , upper_b = third_q + (1.5 * iqr)
  ) %>% cbind (
    low_outliers = c(
      pcauchy(.$lower_b[1])
      , pchisq(.$lower_b[2], df = 1)
      , pexp(.$lower_b[3], rate = 1)
      , pf(.$lower_b[4], df1 = 5, df2 = 2)
      , pt(.$lower_b[5], df = 5)
    ), 
    high_outliers = c(
      1 - pcauchy(.$upper_b[1])
      , 1 - pchisq(.$upper_b[2], df = 1)
      , 1 - pexp(.$upper_b[3], rate = 1)
      , 1 - pf(.$upper_b[4], df1 = 5, df2 = 2)
      , 1 - pt(.$upper_b[5], df = 5)
    )
  )
```


```{r, echo = F}
rownames(boxplot_values) = c("Cauchy", "Chi-square", "Exponential", "F", "Student")
boxplot_values %>% 
  mutate(across(where(is.numeric), round, 4)) %>%
  kbl(
    format = "html"
    , caption = "Theorical Distribution values for Boxplot Definition"
    , col.names = c(
      "First Quartile"
      , "Second Quartile"
      , "Third Quartile"
      , "IQR"
      , "Lower Bound"
      , "Upper Bound"
      , "Prob. Low Outliers"
      , "Prob. High Outliers")
  ) %>%
    kable_styling()
```

# Empirical Values

Generate 10.000 random observation for each distribution and compare the empirical results to the theoretical values expected. 

```{r, fig.cap="Boxplot for 10.000 Random values for each Distribution", fig.align='center', warning=F, message=F}
set.seed(45)
n = 10000

simulations = data.frame (
  cauchy_sim = rcauchy(n)
  , chisqr_sim = rchisq(n, df = 1)
  , exp_sim = rexp(n, rate = 1)
  , fdist_sim = rf(n, df1 = 5, df2 = 2)
  , student_sim = rt(n, df = 5)
 ) 

simulations %>% 
  rename(
    "Cauchy" = cauchy_sim
    , "Chi-square" = chisqr_sim
    , "Exponential" = exp_sim
    , "F" = fdist_sim
    , "Student" = student_sim
  ) %>% 
  melt() %>% 
  ggplot(
    aes(x = value, col = variable) 
  ) + 
  geom_boxplot() + 
  facet_wrap(~variable, scale = "free") 
```


```{r}
sim_summary = apply(
  simulations
  , MARGIN = 2
  , FUN = quantile
) %>% t()

emp_values = sim_summary[,c(2,3,4)] %>% 
  data.frame() %>% 
  rename(first_q = X25., second_q = X50., third_q = X75.) %>% 
  mutate(
    iqr = third_q - first_q
    , lower_b = first_q - 1.5 * iqr
    , upper_b = third_q + 1.5 * iqr
    , low_outliers = colSums(sweep(simulations, 2, lower_b, `<`))/nrow(simulations)
    , high_outliers = colSums(sweep(simulations, 2, upper_b, `>`))/nrow(simulations)
  )


```

```{r, echo = F}
rownames(emp_values) = c("Cauchy", "Chi-square", "Exponential", "F", "Student")
emp_values %>% 
  mutate(across(where(is.numeric), round, 4)) %>%
  kbl(
    format = "html"
    , caption = "Empirical values for Boxplot Definition"
    , col.names = c(
      "First Quartile"
      , "Second Quartile"
      , "Third Quartile"
      , "IQR"
      , "Lower Bound"
      , "Upper Bound"
      , "Prob. Low Outliers"
      , "Prob. High Outliers")
  ) %>%
    kable_styling()
```

# Comparison Theoretical vs Empirical

```{r}
difference = t(emp_values) - t(boxplot_values)
```


```{r, echo = F}
rownames(difference) = c(
      "First Quartile"
      , "Second Quartile"
      , "Third Quartile"
      , "IQR"
      , "Lower Bound"
      , "Upper Bound"
      , "Prob. Low Outliers"
      , "Prob. High Outliers")

difference %>% 
  data.frame() %>%
  mutate(across(where(is.numeric), round, 4)) %>%
  kbl(
    format = "html"
    , caption = "Differences Between Empirical and Theoretical Values"
    , col.names = c(
      "Cauchy"
      , "Chi-square"
      , "Exponential"
      , "F"
      , "Student"
      )
  ) %>%
    kable_styling()
```

